---
title: "DA5020.A6.Bhavitha.Kandru"
author: "Bhavitha Kandru"
date: "2023-10-18"
output: html_document
---

```{r}
library(tidyverse)
library(DBI)
library(RSQLite)
library(ggplot2)
library(scales)
```

```{r}
mydb<- dbConnect(SQLite(), "/Users/bhavithakandru/Desktop/6020/data/imdb.db")
dbListTables(mydb)
```

# What is the average runtime for the thriller movie genre.
```{r}
sqlcmd="SELECT * FROM movie_info"
movie=dbGetQuery(mydb, sqlcmd)

query <- "SELECT AVG(CAST(SUBSTR(Runtime, 1, LENGTH(Runtime) - 4) AS REAL)) AS Average_Runtime
          FROM movie_info
          WHERE Genre LIKE '%Thriller%'"

result <- dbGetQuery(mydb, query)

cat("Average Runtime for Thriller Movies:", result$Average_Runtime, "minutes\n")
```

#Question 1(1) : Create a table named director_info using SQLite; the columns are: Director_ID, and Director_Name. The Director_ID should be the primary key.
#         1(2) : Import the entire data from the CSV file into the director_info table using the SQLite .import command (see helpful resources below). Verify that your data was imported correctly.
```{r}
# .open "/Users/bhavithakandru/Desktop/6020/data/imdb.db" 
# create table director_info(Director_ID int PRIMARY KEY, Director_Name string)
# .separator ',' 
# .import /Users/bhavithakandru/Desktop/6020/data/directors.csv director_info
dbListFields(mydb,"director_info")
sqlcmd="SELECT * FROM director_info"
director=dbGetQuery(mydb,sqlcmd)
```

#Question 2(1) : Count the number of rows in the movie_info and director_info tables.
```{r}
movie_count <- dbGetQuery(mydb, "SELECT COUNT(*) FROM movie_info")
director_count <- dbGetQuery(mydb, "SELECT COUNT(*) FROM director_info")
movie_count
director_count
```

#Question 2 (2) : How many movies were released between 2010 and 2020 (inclusive)? Visualize the results.
```{r}
query2 <- "SELECT Release_Year, COUNT(*) AS MovieCount
          FROM movie_info
          WHERE Release_Year >= '2010' AND Release_Year <= '2020'
          GROUP BY Release_Year
          ORDER BY Release_Year"
movie_release <- dbGetQuery(mydb, query2)
print(movie_release)

#Visualization
ggplot(movie_release, aes(x = Release_Year, y = MovieCount)) +
  geom_bar(stat = 'identity', fill = 'red') +
   geom_text(aes(label = MovieCount), vjust = -0.5, size = 3) +  # Add movie count labels
  labs(title = "Number of Movies Released Each Year (2010-2020)", x = "Year", y = "Number of Movies")
```

#Question 2 (3) : What is the minimum, average and maximum ratings for “Action” movies. Ensure that you query the genre using wild cards.

```{r}
query3 <- "SELECT MIN(IMDB_Rating) AS Min_Rating, AVG(IMDB_Rating) AS Avg_Rating, MAX(IMDB_Rating) AS Max_Rating FROM movie_info WHERE Genre LIKE '%Action%'"
ratings <- dbGetQuery(mydb, query3)
ratings
```

#Question 2 (4) : What are the 25 highest-grossing movies within the dataset? Display the title, genre and gross.

```{r}
query4 <- "SELECT Series_Title, Genre, CAST(Gross AS REAL) AS Gross FROM movie_info ORDER BY Gross DESC LIMIT 25"
top_grossing_movies <- dbGetQuery(mydb, query4)
top_grossing_movies
```

#Question 2 (5) : Which directors have the highest-grossing movies. Display the director name and the total gross. Ensure that you join the necessary tables. Visualize the results using a Bar chart.

```{r}
query5 <- "SELECT d.Director_Name, SUM(m.Gross) AS Total_Gross
          FROM director_info AS d
          JOIN movie_info AS m ON d.Director_ID = m.Director_ID
          GROUP BY d.Director_Name
          ORDER BY Total_Gross DESC
          LIMIT 10"
top_directors <- dbGetQuery(mydb, query5)

#Visualization
ggplot(data = top_directors, aes(x = reorder(Director_Name, Total_Gross), y = Total_Gross)) +
  geom_bar(stat = "identity", fill = "orange") +
  labs(title = "Top Directors by Total Gross", x = "Director Name", y = "Total Gross") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
#Question 2 (6) : Create a function called verifyDirector() that takes a director name as its argument, and queries the database to check if the director exists. Your function should display a message to notify the user if the director was found or not.

```{r}
verifyDirector <- function(director_name) {
  query6 <- sprintf("SELECT COUNT(*) FROM director_info WHERE Director_Name = '%s'", director_name)
  result <- dbGetQuery(mydb, query6)$'COUNT(*)'
  if (result > 0) {
    cat(paste("Director", director_name, "was found in the database.\n"))
  } else {
    cat(paste("Director", director_name, "was not found in the database.\n"))
  }
}

# Use the function to verify a director (replace 'DirectorName' with the actual name)
verifyDirector("David Yates")
```

