---
title: "DA5020.A1.Bhavitha.Kandru"
author: "Bhavitha Naga Sai Kandru"
date: "2023-09-09"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Question1(2):
#Dataframe: A dataframe is a tabular format which is used to store and organize data, where each coulmn contains different types of data and each row has an individual record. My example dataframe is:
```{r}
players <- c('Sachin Tendulkar','Virat Kohli','Rohit Sharma','Ricky Ponting','AB de Villiers','Chris Galye')
cent <- c(49,46,30,30,25,25)
cricket<-data.frame(players,cent)
cricket
```

#Question 2:
```{r cars}
#Creating variables
cars<-c('Truck','Car','SUV')
mpg<-c(11, 30, 24)
cost<-c(5000,25000,35000)
vehicles<-data.frame(cars,mpg,cost)
#Loading Data
data(vehicles)
#Printing Data
print(vehicles)
```

#Question 2a:
```{r}
vehicles<-data.frame(cars,mpg,cost)
vehicles[1,3]
```
#The cost of the truck was displayed which is 5000 

```{r}
vehicles<-data.frame(cars,mpg,cost)
vehicles[1:3]
```
# The rows from 1 to 3 are displayed

```{r pressure, echo=FALSE}
vehicles<-data.frame(cars,mpg,cost)
vehicles[,3]
```
# The cost of all the vechicles was displayed.

#question3:
```{r}
data(mtcars)
head(mtcars,3)
```

```{r}
data(mtcars)
tail(mtcars,5)
```
#Question4
```{r}
#Categorical Variable
category <- names(sapply(mtcars,typeof))
str(category[c(2,8,9,10,11)])
```
```{r}
# Continuous variable
continuous <- names(sapply(mtcars,typeof))
str(continuous[c(1,3,4,5,6,7)])
```

#Question5
```{r}
library(ggplot2)
ggplot(mtcars,aes(x=disp,y=mpg)) 
```
#The given code do not display any relationship between 'disp' and 'mpg' because it only excute the plot but not specifying the layers.

```{r}
ggplot(mtcars,aes(x=disp,y=mpg)) + geom_point()
```
#To get the relation between 'disp' and 'mpg',add geom_point(). It shows a scatter plot between 'disp' and 'mpg'.

#Question6
```{r}
data(mpg)
ggplot(mpg,aes(x=displ,y=hwy,color=cyl))+geom_point()
```
#As the number of cylinders increases the highway per gallon(hwy) reduces and also with increase in cylinders the engine displacement also increases.

#Question7
```{r}
ggplot(mpg,aes(x=displ,y=hwy,color=cyl))+geom_point() +facet_wrap(vars(drv))
```
#When compared to four-wheel drive (4WD) and front-wheel drive (FWD), rear-wheel drive (RWD) vehicles have a higher engine displacement but fewer cylinders, ranging from 3 to 7 liters. In contrast, 4WD and FWD vehicles exhibit nearly identical highway miles per gallon (MPG) but possess smaller engine displacements.
